Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
revised Intro

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are often genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].

. To understand how this variation of thermal tolerance evolved, we need information on the diversity of genes, and their expression patterns, involved in thermal tolerance.

In this study, we present a novel method for gene expression analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient, and report the set of genes that are thermally-responsive; the *thermal reactionome*. We then use a comparative approach to compare the thermal reactionomes between two closely-related ant species to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.

@@

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species. For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A. carolinensis*, or colder temperatures in *A. picea*. Additionally, genes may change their sensitivity to thermal extremes. Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

@@

## Why ants?

>>>>

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffman2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their expression patterns, involved in thermal tolerance.

@@

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??].

Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways.

First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@...] and other protective genes to a greater extent than *A. carolinensis* , while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies upon proteome stability [@...] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

Second, responsive genes may change their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero...]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes.

Moreover, insight on the evolution of thermal tolerance can be gained by examining shifts in the patterns of responsive genes between species. For example, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

@@
